A Study of Common Bug Fix Patterns in Rust \thanks{Competing Interests: We have no competing interests and are funded by a Discovery Grant from Canada's Natural Science and Engineering Research Council.}

### 2023-06-04 discussion.tex
resolved all but 3 of the todos

\todo{this next paragraph doesn't seem to fit here, do we need it?}
A general approach for code embedding is to extract paths with semantic information that better serve a given embedding goal. For instance, to design an embedding method for repairing Rust programs, the designer might want to disregard certain elements, such as moving ownership to closure bodies. Such elements may be irrelevant for repair because they did not appear in the bug patterns that we presented.

>>>>

% A general approach for code embedding is to extract paths with semantic information that better serve a given embedding goal. For instance, to design an embedding method for repairing Rust programs, the designer might want to disregard certain elements, such as moving ownership to closure bodies. Such elements may be irrelevant for repair because they did not appear in the bug patterns that we presented.

### 2023-06-04 introduction.tex
ready for resubmit
reviewer2 - minor revision#1 - put Rust intro in the beginning
add two todos
resolved all but 3 of the todos

@@

Many studies have investigated detecting and categorizing bug fix patterns in general~\citep{islam2021changes,madeiral2018towards,pan2009toward}, or for a specific programming language~\citep{yang2022mining,hanam2016discovering,campos2019discovering}. In this work, we study common bug patterns that appear in Rust projects. Rust is a relatively new programming language~\citep{klabnik2019rust} which allows programmers to write programs that have low-level control over resources, while still ensuring high-level safety guarantees (for programs written in safe Rust). What distinguishes Rust from the other systems programming languages, e.g. C/C++, is its ownership framework, which is an integral part of Rust. The ownership framework keeps track of references to different memory locations, and will free them once they go out of scope. If a program passes the compiler’s checks, it meets several safety properties by construction; e.g. it has no dangling references or double-frees. Such properties had to be manually ensured (or not!) in languages like C/C++. Moreover, not only does ownership remove the need for a garbage collector in the runtime environment, it frees the programmer of the responsibility to ensure certain safety-related properties~\citep{qin2020understanding}. Due to the memory and thread safety guarantees that Rust provides, many software companies have adopted Rust in production. For example, Dropbox decided to migrate to Rust for developing their storage system, since they were not completely satisfied with the features provided by AWS’s S3 or Golang~\citep{dropbox}.

The Rust compiler's Borrow-Checker component enforces ownership rules and hence many of Rust's safety properties. Because of the Borrow-Checker's importance in Rust, we split the patterns that we present into two groups: (1) non-borrow-checker related patterns (General Patterns), and (2) borrow-checker related patterns (BC-Related Patterns). \\

@@

\begin{itemize}
    \item To the best of our knowledge, this work is the first to exploit a code analysis pipeline to automatically mine Rust open source projects and extract common cross-project bug fix patterns.
    \item This work proposes a novel method to encode key AST information in fixed size datapoints using a semi-automatically derived weighting scheme.
    \item Keeping in mind the key role of the borrow checker in Rust, this work specifically identifies bug fix patterns that are related to Rust's ownership framework in addition to general bug fix patterns for Rust.
\end{itemize}

>>>>

Rust is a relatively new programming language~\citep{klabnik2019rust} which allows programmers to write programs that have low-level control over resources, while still ensuring high-level safety guarantees (for programs written in safe Rust). What distinguishes Rust from the other systems programming languages, e.g. C/C++, is its ownership framework, which is an integral part of Rust. The ownership framework keeps track of references to different memory locations, and will free them once they go out of scope. If a program passes the compiler’s checks, it meets several safety properties by construction; e.g. it has no dangling references or double-frees. Such properties had to be manually ensured (or not!) in languages like C/C++. Moreover, not only does ownership remove the need for a garbage collector in the runtime environment, it frees the programmer of the responsibility to ensure certain safety-related properties~\citep{qin2020understanding}. Due to the memory and thread safety guarantees that Rust provides, many software companies have adopted Rust in production. For example, Dropbox decided to migrate to Rust for developing their storage system, since they were not completely satisfied with the features provided by AWS’s S3 or Golang~\citep{dropbox}.

@@

Many studies have investigated detecting and categorizing bug fix patterns in general~\citep{islam2021changes,madeiral2018towards,pan2009toward}, or for a specific programming language~\citep{yang2022mining,hanam2016discovering,campos2019discovering}. In this work, we study common bug patterns that appear in Rust projects. The Rust compiler's Borrow-Checker component enforces ownership rules and hence many of Rust's safety properties. Because of the Borrow-Checker's importance in Rust, we split the patterns that we present into two groups: (1) non-borrow-checker related patterns (General Patterns), and (2) borrow-checker related patterns (BC-Related Patterns). \\

@@

\begin{itemize}
    \item To the best of our knowledge, this work is the first to exploit a code analysis pipeline to automatically mine Rust open source projects and extract common cross-project bug fix patterns.
    \item This work proposes a novel method to encode key AST information in fixed size datapoints using a semi-automatically derived weighting scheme.
    \item Keeping in mind the key role of the borrow checker in Rust, this work specifically identifies bug fix patterns that are related to Rust's ownership framework in addition to general bug fix patterns for Rust.
    \item We identified that the most frequently observed general pattern is the addition or removal of struct fields. Among the BC-related patterns, we found that the most common pattern is the removal of a clone() call. 
\end{itemize}

\noindent \textbf{Data Availability Statement} The datasets generated and analyzed during the current study are available in the Zenodo repository, \url{https://zenodo.org/record/7388618}.

### 2023-06-04 main.tex
resolved all but 3 of the todos

Yet, to the best of our knowledge, no research has sought to identify the most pervasive bug fix patterns within Rust programs. In this project, we introduce Ruxanne, a tool for analyzing and extracting fix patterns in Rust. Ruxanne implements a novel embedding of Rust code into fixed-sized vectors. Using Ruxanne, we mined the top 18 most-starred Rust projects in Github to discover the most common bug fix patterns committed to their repositories. We analyzed 87,726 code changes drawn from 57,214 commits across these 18 projects. After clustering the code changes, and conducting a manual analysis, we discovered 20 groups of cross-project bug fix patterns, which we categorize as (1) general patterns and (2) borrow-checker-related patterns. We describe each of these patterns and their implications to automated program repair.\todo{talk about the most popular pattern}
\keywords{Bug patterns \and Pattern mining \and Bug fix changes \and Rust}
% \PACS{PACS code1 \and PACS code2 \and more}
\subclass{68N01 \and 68N15} %% Computer Science : General and Programming Languages
\end{abstract}

>>>>

Yet, to the best of our knowledge, no research has sought to identify the most pervasive bug fix patterns within Rust programs. In this project, we introduce Ruxanne, a tool for analyzing and extracting fix patterns in Rust. Ruxanne implements a novel embedding of Rust code into fixed-sized vectors. Using Ruxanne, we mined the top 18 most-starred Rust projects in Github to discover the most common bug fix patterns committed to their repositories. We analyzed 87,726 code changes drawn from 57,214 commits across these 18 projects. After clustering the code changes, and conducting a manual analysis, we discovered 20 groups of cross-project bug fix patterns, which we categorize as (1) general patterns and (2) borrow-checker-related patterns. Among the general patterns, the most frequently observed pattern is when the user either adds or removes one of the struct fields. In the case of borrow-checker-related patterns, the most common pattern we encountered is when the user removes a clone() call. We describe all the patterns and their implications to automated program repair.
\keywords{Bug patterns \and Pattern mining \and Bug fix changes \and Rust}
% \PACS{PACS code1 \and PACS code2 \and more}
\subclass{68N01 \and 68N15} %% Computer Science : General and Programming Languages
\end{abstract}

### 2023-06-04 related.tex
add todo about refactoring tools
resolved all but 3 of the todos

>>>>

\subsection{Refactoring tools for Rust}
Clippy is a widely known as the best code refactoring tool for Rust that can identify over 600 common mistakes and offer automated fixes. It provides users with the flexibility to adjust the linting level, allowing them to control the granularity of the lint suggestions they receive. As discussed in Section~\ref{sec:bc_patterns}, we found that Clippy can detect and address five of our borrow-checker-related patterns. \cite{sam2017automated} introduced a refactoring tool that uses the Rust compiler infrastructure to perform various refactoring operations. These operations encompass renaming variables, arguments, fields, functions, methods, structs, and enumerations, as well as inlining local variables, and reifying and eliding lifetime parameters. In a more advanced use case, \cite{ling2022rust} proposed CRustS, a tool designed for refactoring C2Rust output. C2Rust is a C to Rust transpiler that converts C code into Rust syntax but retains the unsafe semantics of C. By relaxing the constraints of preserving semantics during transformations, CRustS increases the proportion of transformed code that successfully passes the safety checks of the Rust compiler.

### 2023-06-04 results.tex
add todo about manual trim
resolved all but 3 of the todos

\begin{tabular}{l|l|r|r}
\textbf{Catgegory} & \textbf{ID} & \textbf{\# Instances} & \textbf{\# Projects} \\
\hline
\multirow{2}{*}{Attributes} & G.attr.struct & 172 & 16                                  \\
& G.attr.struct.field & 62 & 16                     \\\cline{1-4}
\multirow{2}{*}{Struct} & G.struct.field  & 179 & 17\\
& G.struct.field.pub & 23 & 12   
\\\cline{1-4}
Option & G.option & 22 & 12                    \\\cline{1-4}
\multirow{4}{*}{Types} & G.type.field  & 72 & 14 \\
& G.type.generic & 46 & 14\\ & G.type.enum.variant  & 39 & 12 \\
& G.type.tuple & 6 & 6                                         \\\cline{1-4}
\multirow{1}{*}{Traits} & G.trait.bound & 6 & 5 \\\cline{1-4}
\multirow{2}{*}{Match} & G.match.pattern & 41 & 12 \\
& G.match.code & 8 & 5 \\
\end{tabular}
\caption{\label{table:general}Bug Fix Pattern Categorization (General Patterns)}
\end{table}

\subsubsection{Attributes}

The most common bug fix pattern in our corpus modifies attributes. In Rust, attributes enable developers to add metadata to program elements. Some of the attributes---presumably all of the ones involved in bug fixes---affect the semantics of the program; for instance, libraries can generate different code based on the attributes that the developer specifies.

@@

The concept of attributes in Rust is similar to Python's decorators\todo{are python decorator changes important in the Python dataset?} and Java's annotations (though Java annotations may be less likely to have runtime effect; for instance, Java @Override adds additional compile-time constraints). While each language may have its own syntax and specific use cases, the fundamental idea of attaching additional information to code elements is shared across these languages. \\

@@

//G.attr.struct.drop: Removing attributes from a struct
<@\textcolor{red}{--}@> #[derive(Serialize, Deserialize, Clone, PartialEq, Eq, Debug)]
<@\textcolor{green}{++}@> #[derive(Serialize, Deserialize, Clone, PartialEq, Eq)]
pub struct Subset {
    /// Invariant, maintained by `SubsetBuilder`: all `Segment`s have non-zero
    /// length, and no `Segment` has the same count as the one before it.
    segments: Vec<Segment>,
}

@@

\noindent\textbf{Description:} In this pattern, the developer changes the attribute set of a struct. There are three variants of this pattern: Adding attributes to the struct (G.attr.struct.add); Removing attributes from the struct (G.attr.struct.drop); Changing the content of an existing attribute (G.attr.struct.change). These attributes are Outer Attributes as we can find struct-related non terminals within the ASTDiff of the instances of this pattern.

\todo{discussion on how many of the changes are of each variant: over 90\% are modify, but get the actual number}

@@

\noindent\textbf{Description:} In this pattern, the attribute set of a struct field changes. As with \textit{\textbf{G.attr.struct}}, there are also three subtypes of this change pattern. Using these attributes, the developers can control different actions they wish to apply on struct fields, in addition to providing meta-information. For instance, in the code snippet provided above, the developers chose to make Serde (a serialization/deserialization tool for Rust) call a different function (\verb+is_empty+ instead of \verb+Option::is_none+) to determine whether to skip serializing the \verb+attributes+ field. \\

\todo{please confirm: G.attr.struct.field is not a subset of G.attr.struct.}
\todo{we should also get the count of subtypes}

\noindent\textbf{Actionability:} Because attributes have semantic effects, program repair tools can effect repairs by adjusting the set of attributes included. Starting from the observation that so many of our bug fixes change attributes, we believe that IDEs need to be able to propose relevant attribute changes, as witnessed in our corpus. The example in the text about changing Deserialize to ConfigDeserialize is specific to that program, but other changes would apply more generally. \todo{which?} Also, we demonstrated a case where the developer removed a \texttt{Debug} attribute. A novel type of performance program repair tool working with a run-time profiler could identify unnecessary code and remove attributes that generate it.

>>>>

\begin{tabular}{l|l|r|r}
\textbf{Catgegory} & \textbf{ID} & \textbf{\# Instances} & \textbf{\# Projects} \\
\hline
\multirow{2}{*}{Attributes} & G.attr.struct & 81 & 16                                  \\
& G.attr.struct.field & 34 & 9                     \\\cline{1-4}
\multirow{2}{*}{Struct} & G.struct.field  & 179 & 17\\
& G.struct.field.pub & 23 & 12   
\\\cline{1-4}
Option & G.option & 22 & 12                    \\\cline{1-4}
\multirow{4}{*}{Types} & G.type.field  & 72 & 14 \\
& G.type.generic & 46 & 14\\ & G.type.enum.variant  & 39 & 12 \\
& G.type.tuple & 6 & 6                                         \\\cline{1-4}
\multirow{1}{*}{Traits} & G.trait.bound & 6 & 5 \\\cline{1-4}
\multirow{2}{*}{Match} & G.match.pattern & 41 & 12 \\
& G.match.code & 8 & 5 \\
\end{tabular}
\caption{\label{table:general}Bug Fix Pattern Categorization (General Patterns)}
\end{table}

\subsubsection{Attributes}
The most common bug fix pattern in our corpus modifies attributes. In Rust, attributes enable developers to add metadata to program elements. Some of the attributes---presumably all of the ones involved in bug fixes---affect the semantics of the program; for instance, libraries can generate different code based on the attributes that the developer specifies.

@@

The concept of attributes in Rust is similar to Python's decorators and Java's annotations (though Java annotations may be less likely to have runtime effect; for instance, Java @Override adds additional compile-time constraints). While each language may have its own syntax and specific use cases, the fundamental idea of attaching additional information to code elements is shared across these languages. We describe bug patterns that involve modification in Attributes. In the case of Python, according to the findings of \cite{yang2022mining}, no bug patterns involving Python's decorators were reported in their study. \\

@@

// G.attr.struct.drop: Removing attributes from a struct
<@\textcolor{red}{--}@> #[derive(Serialize, Deserialize, Clone, PartialEq, Eq, Debug)]
<@\textcolor{green}{++}@> #[derive(Serialize, Deserialize, Clone, PartialEq, Eq)]
pub struct Subset {
    segments: Vec<Segment>,
}

@@

\noindent\textbf{Description:} In this pattern, the developer changes the attribute set of a struct. There are three variants of this pattern: Adding attributes to the struct (G.attr.struct.add); Removing attributes from the struct (G.attr.struct.drop); Changing the content of an existing attribute (G.attr.struct.change). These variants account for 58, 10, and 13 out of 81 datapoints in the cluster, respectively. Thus, adding attributes is the most frequent variant. Alse, these attributes are Outer Attributes as we can find struct-related non terminals within the ASTDiff of the instances of this pattern.

@@

\noindent\textbf{Description:} In this pattern, the attribute set of a struct field changes. As with \textit{\textbf{G.attr.struct}}, there are also three subtypes of this change pattern. Out of 34 datapoints in this cluster, 32 involve changes to a field attribute, while one datapoint each corresponds to the addition and removal of a field attribute. Using these attributes, the developers can control different actions they wish to apply on struct fields, in addition to providing meta-information. For instance, in the code snippet provided above, the developers chose to make Serde (a serialization/deserialization tool for Rust) call a different function (\verb+is_empty+ instead of \verb+Option::is_none+) to determine whether to skip serializing the \verb+attributes+ field. Note that \textit{\textbf{G.attr.struct.field}} is not a subset of \textit{\textbf{G.attr.struct}}. \\

\noindent\textbf{Actionability:} Because attributes have semantic effects, program repair tools can effect repairs by adjusting the set of attributes included. Starting from the observation that so many of our bug fixes change attributes, we believe that IDEs need to be able to propose relevant attribute changes, as witnessed in our corpus. The example for \textit{\textbf{G.attr.struct.change}} about changing \verb+Deserialize+ to \verb+ConfigDeserialize+ is specific to that program, but other changes would apply more generally. Also, as for \textit{\textbf{G.attr.struct.drop}}, we demonstrated a case where the developer removed a \verb+Debug+ attribute. A novel type of performance program repair tool working with a run-time profiler could identify unnecessary code and remove attributes that generate it.
